---
title: "FINAL PROJECT"
author: "Oanh"
date: "2023-08-22"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Covid 19 Data Report

## Introduction:

To begin, we need to install these necessary packages:(tidyverse),(lubridate), (ggplot2),(dplyr),(knitr)

```{r , message=FALSE}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(knitr)
library(dplyr)
```

Read the data from the link.

```{r}
url_in<-"https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names<-c("time_series_covid19_confirmed_global.csv","time_series_covid19_deaths_global.csv","time_series_covid19_confirmed_US.csv","time_series_covid19_deaths_US.csv")
urls<-str_c(url_in, file_names)
global_cases<-read_csv(urls[1])
global_deaths<-read_csv(urls[2])
US_cases<-read_csv(urls[3])
US_deaths<-read_csv(urls[4])
head(global_cases)
```

## Data Preparation and Cleaning

After looking at global_cases and global_deaths, I would like to tidy those datasets and put each variable (date, cases, deaths) in their own column. Also, I don't need Lat and Long for the analysis I am planning, so I will get rid of those and rename Region and State to be more R friendly.

```{r}
global_cases<-global_cases %>% pivot_longer(cols = -c('Province/State','Country/Region',Lat, Long),names_to = "date", values_to = "cases") %>% select(-c(Lat,Long))
head(global_cases,10)
global_deaths<-global_deaths %>% pivot_longer(cols = -c('Province/State','Country/Region',Lat, Long),names_to = "date", values_to = "deaths") %>% select(-c(Lat,Long))
```

Combine cases in to deaths per date into one variable we will call global and rename our country region to get rid of slash mark and the same with province sate.

```{r}
global<-global_cases %>% full_join(global_deaths) %>% rename(Country_Region='Country/Region',Province_State ='Province/State')%>% mutate(date =mdy(date))
```

```{r}
summary(global)
```

Filter out and keep only where the cases are positive.

```{r}
global<-global %>% filter(cases >0)
summary(global)
```

Check the maximum is a valid maximum or if it were a typo.
```{r}
global %>% filter(cases >100000000)
```
We do the same with US_cases and US_deaths and combine cases in to deaths per date into one variable we will call US.
```{r}
US_cases<-US_cases %>% pivot_longer(cols=-(UID:Combined_Key),names_to = "date", values_to = "cases") %>% select(Admin2:cases)%>% mutate(date=mdy(date)) %>% select(-c(Lat, Long_))
US_deaths<-US_deaths %>% pivot_longer(cols=-(UID:Population),names_to = "date", values_to = "deaths") %>% select(Admin2:deaths)%>% mutate(date=mdy(date)) %>% select(-c(Lat, Long_))
```

```{r}
US<-US_cases %>% full_join(US_deaths)
```
```{r}
global<- global %>% unite("Combined_Key", c(Province_State, Country_Region), sep=", ",na.rm=TRUE,remove=FALSE)
global
```
Add population into global.
```{r}
uid_lookup_url<-"https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
uid<-read_csv(uid_lookup_url) %>% select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

global<-global %>% left_join(uid, by=c("Province_State", "Country_Region")) %>% select(-c(UID,FIPS)) %>% select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key)
global
```
## Visualize US by sate
```{r}
US_by_state<- US %>% group_by(Province_State, Country_Region, date) %>% summarise(cases=sum(cases), deaths=sum(deaths), Population=sum(Population)) %>% mutate(deaths_per_mill=deaths*1000000/Population) %>% select(Province_State, Country_Region, date, cases, deaths, deaths_per_mill, Population) %>% ungroup()
US_by_state
```
```{r}
US_totals <-US_by_state %>% group_by(Country_Region, date) %>% summarise(cases=sum(cases), deaths=sum(deaths), Population=sum(Population)) %>% mutate(deaths_per_mill=deaths*1000000/Population) %>% select(Country_Region, date, cases, deaths, deaths_per_mill, Population) %>% ungroup()
US_totals
```
Make plot
```{r}
US_totals %>% filter(cases>0) %>% ggplot(aes(x=date, y = cases)) +geom_line(aes(color="cases"))+geom_point(aes(color="cases")) + geom_line(aes(y=deaths, color="deaths"))+geom_point(aes(y=deaths, color="deaths"))+scale_y_log10()+theme(legend.position = "bottom",axis.text.x = element_text(angle=90))+labs(title="COVID 19 in US",y=NULL)
```                                                                                                                                                                                     
I will do the same plot for New York State
```{r}
state<-"New York"
```
```{r}
US_by_state %>% filter(Province_State==state) %>% filter(cases>0) %>% ggplot(aes(x=date, y = cases)) +geom_line(aes(color="cases"))+geom_point(aes(color="cases")) + geom_line(aes(y=deaths, color="deaths"))+geom_point(aes(y=deaths, color="deaths"))+scale_y_log10()+theme(legend.position = "bottom",axis.text.x = element_text(angle=90))+labs(title=str_c("COVID 19 in ",state),y=NULL)
```                                                                                                                                                                                                                                                                                                                                
## Analyzing about no new cases
First trasform our data again by adding new_cases and new_deaths variables
```{r}
US_by_state<-US_by_state %>% mutate(new_cases=cases-lag(cases), new_deaths=deaths-lag(deaths))
US_totals<-US_totals%>% mutate(new_cases=cases-lag(cases), new_deaths=deaths-lag(deaths))
tail(US_totals,10)
```
  Make plot 
```{r}
US_totals %>% ggplot(aes(x=date, y = new_cases)) +geom_line(aes(color="new_cases"))+geom_point(aes(color="new_cases")) + geom_line(aes(y=new_deaths, color="new_deaths"))+geom_point(aes(y=new_deaths, color="new_deaths"))+scale_y_log10()+theme(legend.position = "bottom",axis.text.x = element_text(angle=90))+labs(title="COVID19 in US ",y=NULL)
```
I will do the same plot for New York State
```{r}
state<-"New York"
US_by_state %>%filter(Province_State==state)%>% ggplot(aes(x=date, y = new_cases)) +geom_line(aes(color="new_cases"))+geom_point(aes(color="new_cases")) + geom_line(aes(y=new_deaths, color="new_deaths"))+geom_point(aes(y=new_deaths, color="new_deaths"))+scale_y_log10()+theme(legend.position = "bottom",axis.text.x = element_text(angle=90))+labs(title=str_c("COVID19 in ",state),y=NULL)
```
## Analyzing the worst and the best state
```{r}
US_state_totals<-US_by_state %>% group_by(Province_State) %>% summarise(deaths=max(deaths), cases=max(cases),population=max(Population), cases_per_thou=1000*cases/population,deaths_per_thou=1000*deaths/population)%>%filter(cases>0,population>0)
US_state_totals %>% slice_min(deaths_per_thou, n=10)
```
The best state is American Samoa
```{r}
US_state_totals %>% slice_max(deaths_per_thou, n=10)
```
The worst state is Arizona

##Modeling 

```{r}
mod<-lm(deaths_per_thou ~ cases_per_thou, data=US_state_totals)
summary(mod)
```
Make plot
```{r}
US_tot_w_pred<-US_state_totals %>% mutate(pred=predict(mod))
US_tot_w_pred %>% ggplot()+geom_point(aes(x=cases_per_thou, y =deaths_per_thou), color="blue")+geom_point(aes(x=cases_per_thou, y =pred), color="red")
```

Based on the linear regression results, we can conclude that there is a positive relationship between the number of cases per thousand and the number of deaths per thousand. In other words, as the number of cases per thousand increases, the number of  deaths per thousand increases also . The R-squared value of 0.2933 indicates that the model explains approximately 29% of the variability in the number of deaths per thoudand. However, it's important to note that correlation does not imply causation, and there may be other factors that contribute to the number deaths beyond just number of cases .
